Forecasting interrupted time series

### 2024-02-18 fits.qmd
Polishing the wording

Time series forecasting models use historical data to estimate future values [@fildes2008forecasting] based on consistent patterns observed in the past. However, time series are sometimes disrupted by unusual events that jeopardise the regularity of the time series pattern. This can be defined as an *interrupted time series*. Time series data may be disrupted by a variety of reasons, ranging from temporary policy changes to natural disasters. Outbreaks, epidemics, and pandemics, for example, may have a considerable impact on population behavior, causing discrepancies and interruptions of data in a variety of industries such as healthcare, pharmaceuticals, transportation, retail, tourism, and traffic management. Similarly, supply chain disruptions induced by factors like strikes or shortages cause irregularities in sales, inventory levels, production, and delivery, often resulting in data gaps. Transportation-related data may be interrupted due to equipment failures, such as malfunctions or breakdowns in vehicles, aircraft, or rail systems, resulting in missing or inaccurate data points. Furthermore, cybersecurity breaches data quality by possibly skewing readings in datasets relating to financial transactions, network traffic, and user activity, complicating time series analysis.

@@

It is important to note that in this study, we consider the problem of forecasting interrupted time series both during and after the event. This is a separate problem from change point detection [@truong2020selective] and anomaly detection [@talagala2020anomaly]. In the situations we investigate, we know that a change has occurred that causes unusual data points and want to forecast what will happen next. Change point detection refers to determining *when* a change occurred [@blazquez2021review], and anomaly detection aims to identify unusual observations with the intention of either minimising their value, or paying particular attention to them, and doing a root-cause investigation. Our study focus is also slightly differs from *intervention modeling*, also refereed to as *interrupted time series analysis and modelling* [@mcdowall2019interrupted], which is used when we have data about an outcome over time and the aim is to assess the impact of an intervention, policy, or program adopted for the whole population at a specific time point. This is a special case of what we are looking at at this paper.

To our knowledge, this is the first study to offer strategies for forecasting interrupted time series, such as COVID-19 affected data. The study proposes and evaluates multiple forecasting strategies designed to handle interruptions in time series forecasting caused by disruptive events. These approaches include a range of models capable of capturing different types of changes, from simple level shifts to more complex alterations in seasonal patterns and trends. By comparing the performance of these models on two real data sets affected by COVID-19 pandemic, the paper offers valuable insights into the effectiveness and practical implications of each approach.The paper offers practical guidance for practitioners seeking to improve the accuracy and reliability of time series forecasts in the presence of disruptive events. By discussing the advantages and disadvantages of each strategy and providing recommendations based on the findings, the paper equips practitioners with actionable insights for addressing the challenges associated with forecasting during and after disruptive events. Furthermore, by adhering to reproducibility principles and providing both data and R codes for the forecasting models, the research not only contributes to transparency but also promotes the generalizability of the suggested strategies to a variety of domains. This openness allows for the replication of results, ensuring the reliability of findings, and enhancing the accessibility and utility of study outputs for the larger research community.

@@

One modelling approach that has gained popularity is the intervention model, also known as interrupted time series (ITS) modelling, introduced @box1975intervention. This approach focuses on understanding how and whether outcomes change following the implementation of an intervention, policy, or programme at a specific time point. ITS models provide a robust tool for analysing policy and programme evaluation across diverse domains [@bernal2017interrupted;@mcdowall2019interrupted]. These models can forecast future values by developing models that incorporate a range of components, such as pre-intervention levels, trends, seasonality, covariates related to the intervention time, and other external factors that might affect the outcome variable. Another related approach is the multiple regimes model, which is particularly useful in forecasting due to its ability to capture structural changes and transitions observed in interrupted time series data [@Koopg2007].

Statistical models developed to detect changes in time series data and handle unusual observations have also proven beneficial in similar circumstances. Methods for detecting changing points in time series data may include simple location shift and scale shift models[@aue2013structural], Bayesian [@barry1993bayesian], and supervised and supervised machine learning [@aminikhanghahi2017survey]. Anomaly detection seeks to find points or subsequences in time series data that have unexpected or anomalous values. Facebook's Prophet is an example of a time series forecasting system that uses changepoint detection and automatically adjusts model parameters to catch the most recent trends in a time series.

@@

In the context of forecasting disruptions like the COVID-19 pandemic, Bayesian approaches have been proposed for detecting changing points in outbreaks and forecasting case numbers under counterfactual scenarios [@dehning2020inferring]. Counterfactual analysis examines what would have happened to the forecast variable of interest if the disruptive event had not occurred. In line with this, @athanasopoulos2023probabilistic proposed a variant of this approach for forecasting tourism recovery from COVID-19, combining forecast reconciliation and forecast combinations applied to historical data to generate robust COVID-free counterfactual forecasts that set the expected future paths as if the pandemic never occurred. Comparing those estimations to produced probabilistic forecasts allowed the tourism sector to estimate the difference between scenario-based probabilistic forecasts and forecasts prepared as if the pandemic had never occurred.

@@

An alternative would be to take the model estimated under the previous solution (setting the observations during the disruption to missing) and use it to estimate what might have been during the disruption. Then the model can be re-estimated using the adjusted data. Note that forecasts obtained in this way during the disruption will not be true forecasts because they will have used data from the future in computing the adjusted data during the disruption. But post-disruption forecasts will be true forecasts and should be almost the same as those obtained using the previous solution.

@@

The Australian tourism data set is a monthly time series showing the number of short-term overseas visitors to Australia. The data [@tourismdata] are available from January 2000 to May 2023, and are shown in @fig-tourism-plot1. As the borders closed in March 2020, the number of visitors to Australia dropped to near zero and remained there until towards the end of 2021. The borders officially reopened on February 21, 2022, although it seems visitors began to arrive earlier than that.

@@

We have also shown the periods of lockdown due to COVID-19. In Melbourne, there were six separate lockdowns ranging from `r as.numeric(min(lockdowns$Duration))` days to `r as.numeric(max(lockdowns$Duration))` days in length. The first official lockdown period was from March 31, 2020, to `r format(min(lockdowns$End), "%d %b %Y")`, and the last lockdown was from `r format(max(lockdowns$Start), "%d %b %Y")` to `r format(max(lockdowns$End), "%d %b %Y")`. However, the first period was preceded by over a week when most people elected to stay and work at home, so we have chosen to start the first period on `r format(min(lockdowns$Start), "%d %b %Y")`, to better reflect human behaviour. Otherwise, we have used the official lockdown dates [@MelbourneLockdowns].

@@

We have not previously used the weighted approach in practice, and software to implement it is not currently available. However, it seems to be a good compromise between the highly adaptive model and the intervention model, and we expect it to be useful in many situations once software is available to implement it easily.

@@

It is not uncommon for modellers to experience interruptions in time series data. Whether you're in charge of developing forecasting models or a manager who relies on forecasts to inform decisions and policies, you've undoubtedly recently faced the issue of effectively forecasting time series data influenced by the COVID-19 pandemic. While this can stand as a critical challenge for many forecasting teams, it is important to recognise that similar challenges could arise in circumstances other than COVID-19. Time series can be disrupted by various factors, such as natural disasters, policy changes, price fluctuations, definition changes, sensor failures in the IoT (Internet of Things), maintenance, and breakdowns in factories,  just to mention a few.

The aim of this study is not to compile a comprehensive list of forecasting models applicable to interrupted time series data. Rather, we intend to propose overarching strategies to address such situations, with each strategy allowing for a variety of forecasting methodologies. As we provide publicly accessible R codes, readers are encouraged to adapt the study by employing other forecasting methods tailored to their specific contexts.

@@

When determining a suitable approach to forecasting interrupted time series, several factors should be considered. To begin, the complexity of the interruption itself is critical to approach selection. Intervention models are a suitable approach for interruptions with known features or intricate patterns since they can represent the disruption. In contrast, highly adaptable models give a simpler option for fast adaptations to interruptions without explicit modelling, which is especially useful when the time or effect of the interruption is unknown. Balancing simplicity and flexibility is critical to ensuring that the selected technique is consistent with the features of the interruption and the required forecasting accuracy. Furthermore, the width of prediction intervals should be carefully assessed to determine forecast uncertainty, particularly during and after interruptions. Finally, for greater robustness and accuracy, investigating ensemble models that integrate various forecasting methods might be beneficial.

@@

To enhance reproducibility and facilitate the adoption of proposed strategies, we provide the data and R codes, as well as the entire paper written in R using [Quarto](https://quarto.org/) and the targets R package [@targetsr2021]. All materials required to reproduce this paper will be accessible via a public GitHub repository once the paper is accepted for publication.

>>>>

Time series forecasting models use historical data to estimate future values [@fildes2008forecasting] based on consistent patterns observed in the past. However, time series are sometimes disrupted by unusual events that jeopardise the regularity of the time series pattern. This can be defined as an *interrupted time series*. Time series data may be disrupted by a variety of reasons, ranging from temporary policy changes to natural disasters. Outbreaks, epidemics, and pandemics, for example, may have a considerable impact on population behavior, causing discrepancies and interruptions of data in a variety of industries such as healthcare, pharmaceuticals, transportation, retail, tourism, and traffic management. Similarly, supply chain disruptions induced by factors such as strikes or shortages cause irregularities in sales, inventory levels, production, and delivery, often resulting in data gaps. Transportation-related data may be interrupted due to equipment failures, such as malfunctions or breakdowns in vehicles, aircraft, or rail systems, resulting in missing or inaccurate data points. Furthermore, cybersecurity breaches data quality by possibly skewing readings in datasets relating to financial transactions, network traffic, and user activity, complicating time series analysis.

@@

It is important to note that in this study, we consider the problem of forecasting interrupted time series both during and after the event. This is a separate problem from change point detection [@truong2020selective] and anomaly detection [@talagala2020anomaly]. In the situations we investigate, we know that a change has occurred which has caused some unusual observations, and we want to forecast what will happen next. Change point detection refers to determining *when* a change occurred [@blazquez2021review], and anomaly detection aims to identify unusual observations with the intention of either minimising their value, or paying particular attention to them, and doing a root-cause investigation. Our study focus also differs from *intervention modeling*, sometimes referred to as *interrupted time series analysis and modelling* [@mcdowall2019interrupted], which is used when we have data about an outcome over time and the aim is to assess the impact of an intervention, policy, or program implemented at a specific time point. We are not trying to measure the impact of the interruption; rather we are trying to produce sensible forecasts during and after the interruption.

To our knowledge, this is the first study to describe and compare general strategies for forecasting interrupted time series, such as COVID-19 affected data. The study proposes and evaluates multiple forecasting strategies designed to handle interruptions in time series forecasting caused by disruptive events. These approaches include a range of models capable of capturing different types of changes, from simple level shifts to more complex alterations in seasonal patterns and trends. By comparing the performance of these models on two real data sets affected by the COVID-19 pandemic, the paper offers valuable insights into the effectiveness and practical implications of each approach. By discussing the advantages and disadvantages of each strategy, and providing recommendations based on the findings, the paper equips practitioners with actionable insights for addressing the challenges associated with forecasting during and after disruptive events. Further, by adhering to reproducibility principles and providing both data and R code for the forecasting models, the research contributes to transparency and promotes the generalizability of the suggested strategies to a variety of domains. This openness allows for the replication of results, ensuring the reliability of findings, and enhancing the accessibility and utility of study outputs for the larger research community.

@@

One modelling approach that has gained popularity is the intervention model, also known as interrupted time series (ITS) modelling, introduced by @box1975intervention. This approach focuses on understanding how and whether outcomes change following the implementation of an intervention, policy, or programme at a specific time point. ITS models provide a robust tool for analysing policy and programme evaluation across diverse domains [@bernal2017interrupted;@mcdowall2019interrupted]. These models can forecast future values by developing models that incorporate a range of components, such as pre-intervention levels, trends, seasonality, covariates related to the intervention time, and other external factors that might affect the outcome variable. A related approach is the multiple regimes model, which is particularly useful in forecasting due to its ability to capture structural changes and transitions observed in interrupted time series data [@Koopg2007].

@@

In the context of forecasting disruptions such as the COVID-19 pandemic, Bayesian approaches have been proposed for detecting changing points in outbreaks and forecasting case numbers under counterfactual scenarios [@dehning2020inferring]. Counterfactual analysis examines what would have happened to the forecast variable of interest if the disruptive event had not occurred. @athanasopoulos2023probabilistic proposed a variant of this approach for forecasting tourism recovery from COVID-19, combining forecast reconciliation and forecast combinations applied to historical data, to generate COVID-free counterfactual forecasts of what might have been if the pandemic never occurred. Then scenario-based judgemental probabilistic forecasts were compared with the counterfactual forecasts, to better understand the future recovery of the tourism sector from the pandemic.

@@

An alternative would be to take the model estimated under the previous solution (setting the observations during the disruption to missing), and use it to estimate what might have been during the disruption. Then the model can be re-estimated using the adjusted data. Note that forecasts obtained in this way during the disruption will not be true forecasts because they will have used data from the future in computing the adjusted data during the disruption. But post-disruption forecasts will be true forecasts and should be almost the same as those obtained using the previous solution.

@@

The Australian tourism data set is a monthly time series showing the number of short-term overseas visitors to Australia. The data [@tourismdata] are available from January 2000 to May 2023, and are shown in @fig-tourism-plot1. As the borders closed in March 2020, the number of visitors to Australia dropped to near zero and remained there until towards the end of 2021. The borders officially reopened on 21 February 2022, although it seems visitors began to arrive earlier than that.

@@

We have also shown the periods of lockdown due to COVID-19. In Melbourne, there were six separate lockdowns ranging from `r as.numeric(min(lockdowns$Duration))` days to `r as.numeric(max(lockdowns$Duration))` days in length. The first official lockdown period was from 31 March 2020, to `r format(min(lockdowns$End), "%d %b %Y")`, and the last lockdown was from `r format(max(lockdowns$Start), "%d %b %Y")` to `r format(max(lockdowns$End), "%d %b %Y")`. However, the first period was preceded by over a week when most people elected to stay and work at home, so we have chosen to start the first period on `r format(min(lockdowns$Start), "%d %b %Y")`, to better reflect human behaviour. Otherwise, we have used the official lockdown dates [@MelbourneLockdowns].

@@

We have not previously used the weighted approach in practice, and open-source software to implement it is not currently available. However, it seems to be a good compromise between the highly adaptive model and the intervention model, and we expect it to be useful in many situations once software is available to implement it easily.

@@

It is not uncommon for modellers to experience interruptions in time series data. Both those in charge of developing forecasting models, and managers relying on forecasts to inform decisions and policies, have recently faced the issue of effectively forecasting time series data influenced by the COVID-19 pandemic. While this can stand as a critical challenge for many forecasting teams, it is important to recognise that similar challenges could arise in circumstances other than COVID-19. Time series can be disrupted by various factors, such as natural disasters, policy changes, price fluctuations, definition changes, sensor failures in the IoT (Internet of Things), maintenance, and breakdowns in factories,  just to mention a few.

The aim of this study is not to compile a comprehensive list of forecasting models applicable to interrupted time series data. Rather, we intend to propose overarching strategies to address such situations, with each strategy allowing for a variety of forecasting methodologies. As we provide publicly accessible R code, readers are encouraged to adapt the study by employing other forecasting methods tailored to their specific contexts.

@@

When determining a suitable approach to forecasting interrupted time series, several factors should be considered. To begin, the complexity of the interruption itself is critical to approach selection. Intervention models are a suitable approach for interruptions with known features since they can represent the disruption explicitly. In contrast, highly adaptable models give a simpler option for fast adaptations to interruptions without explicit modelling, which is especially useful when the time or effect of the interruption is unknown. Balancing simplicity and flexibility is critical to ensuring that the selected technique is consistent with the features of the interruption and the required forecasting accuracy. Further, the width of prediction intervals should be carefully assessed to determine forecast uncertainty, particularly during and after interruptions. Finally, for greater robustness and accuracy, investigating ensemble models that integrate various forecasting methods might be beneficial.

@@

To enhance reproducibility and facilitate the adoption of proposed strategies, we provide the data and R code, as well as the entire paper written in R using [Quarto](https://quarto.org/) and the targets package for R [@targetsr2021]. All materials required to reproduce this paper will be accessible via a public GitHub repository once the paper is accepted for publication.
